An efficient composition of bidirectional programs by tupling and lazy updates

### 2019-12-02 paper/2.minbigul.tex
small fixes, change on explanation of bfoldr

minBiGUL inherited from BiGUL also supports transformations $put$ and $get$ which satisfy two above laws. In addition, we restrict adaptive cases of BiGUL on minBiGUL. Then $put$ and $get$ satisfy one more law, \textsc{PutPut} \cite{Foster:2007:CBT:1232420.1232424}, like the following:

@@

The \textsc{PutPut} law means that a source update should overwrite the effect of previous source updates. Due to the satisfaction of three laws, \textsc{GetPut}, \textsc{PutGet} and \textsc{PutPut}, minBiGUL is very well-behaved.

@@

A minBiGUL program may be either a skip of a function or a replacement or a product of two minBiGUL programs or a source/view rearrangement or a case combinator without adaptive cases or a composition of some minBiGUL programs. We use numbers, pairs and lists to construct the program input including the source and/or the view.

@@

In the above definitions, we use if-then-else-fi statements to express semantics of $\putbxinline{Case}$ and $\getbxinline{Case}$ where $\Leftarrow$ means assignments. This statement is useful to describe many functions related to $Case$ in this paper. Statement (if $E_1$ then $X_1$ else $X_2$ fi $E_2$) means if the test $E_1$ is true, the statement $X_1$ is executed and the assertion $E_2$ must be true, otherwise, i.e. $E_2$ is false, the statement $X_2$ is executed and the assertion $E_2$ must be false. If the values of $E_1$ and $E_2$ are distinct, the if-then-else-fi structure is undefined. We can write the equivalent if-then-else statement as follows:

$\tab \text{if } E_1 \text{ then } X_1 \text{ else } X_2 \text{ fi } E_2; S\\
\tabs{2} \ \equiv \text{if }E_1 = true \text{ then } \{ X_1; \text{ if } E_2 = true \text{ then } S \text{ else assert } false; \}\\
    \tabs{4} \text{else } \{ X_2; \text{ if } E_2 = false \text{ then } S \text{ else assert } false; \}$

\noindent We now consider the definition of $phead$ in minBiGUL:\\
$\tab phead = RearrS \ f_1 \ f_2 \ bx_s \ \text{where: } f_1 = \lambda (s::ss).(s,ss), \, f_2 = \lambda (s,ss).(s::ss),\\
    \qtab bx_s = RearrV \ g_1 \ g_2 \ bx_v \ \text{where: } g_1 = \lambda v.(v,()), \, g_2 = \lambda (v,()).v,\\
        \qtabs{2} bx_v = \product{Replace}{(Skip \ (\lambda \_.())}$

The above program rearranges the source, a non-empty list, to a pair of its head element $s$ and its tail $ss$, and the view to a pair $(v, ())$, then we can use $v$ to replace $s$ and $()$ to keep $ss$. Intuitively, $\putbx{phead}{s_0}{v_0}$ returns a list whose head is $v_0$ and tail is the tail of $s_0$, and $\getbx{phead}{s_0}$ returns the head of the list $s_0$. For instance, $\putbx{phead}{[1,2,3]}{100} = [100,2,3]$ and $\getbx{phead}{[1,2,3]} = 1$. If we wanna update the head element of the head element of a list of lists by using the view, we can define a composition like $phead \circ phead$. For example:

@@

In the end of this section, let's see the definition of $bfoldr$ which is a putback function of an important higher-order function on lists, $foldr$:

@@

\noindent $bfoldr$ will look like a composition of many programs at some points in the evaluation if we slow down this process. Since $\circ$ has a higher priority than $\times$, it is seemingly impossible to transform $bfoldr$ from the left associative composition style to the right one.

>>>>

minBiGUL inherited from BiGUL also supports transformations $put$ and $get$ which satisfy two above laws. In addition, we restrict 'adaptive case's of BiGUL on minBiGUL. Then $put$ and $get$ satisfy one more law, \textsc{PutPut} \cite{Foster:2007:CBT:1232420.1232424}, like the following:

@@

The \textsc{PutPut} law means that a source update should overwrite the effect of previous source updates. Due to the satisfaction of three laws, \textsc{GetPut}, \textsc{PutGet} and \textsc{PutPut}, minBiGUL is very well-behaved \cite{Foster:2007:CBT:1232420.1232424}.

@@

A minBiGUL program may be either a skip of a function or a replacement or a product of two minBiGUL programs or a source/view rearrangement or a case combinator without adaptive cases or a composition of some minBiGUL programs. We use numbers, pairs and lists to construct the program inputs including the source and/or the view.

@@

In the above definitions, we use if-then-else-fi statements to express semantics of $\putbxinline{Case}$ and $\getbxinline{Case}$ where $\Leftarrow$ means assignments. This statement is useful to describe many functions related to $Case$ in this paper. Statement (if $E_1$ then $X_1$ else $X_2$ fi $E_2$) means ``if the test $E_1$ is true, the statement $X_1$ is executed and the assertion $E_2$ must be true, otherwise, if $E_2$ is false, the statement $X_2$ is executed and the assertion $E_2$ must be false.'' If the values of $E_1$ and $E_2$ are distinct, the if-then-else-fi structure is undefined. We can write the equivalent if-then-else statement as follows:

$\tab \text{if } E_1 \text{ then } X_1 \text{ else } X_2 \text{ fi } E_2; S\\
\tabs{2} \ \equiv \text{if }E_1 = true \text{ then } \{ X_1; \text{ if } E_2 = true \text{ then } S \text{ else assert } false \}\\
    \tabs{4} \text{else } \{ X_2; \text{ if } E_2 = false \text{ then } S \text{ else assert } false \}$

\noindent As an example of minBiGUL program, consider the definition of $phead$ :\\
$\tab phead = RearrS \ f_1 \ f_2 \ bx_s \ \text{where: } f_1 = \lambda (s::ss).(s,ss), \, f_2 = \lambda (s,ss).(s::ss),\\
    \qtab bx_s = RearrV \ g_1 \ g_2 \ bx_v \ \text{where: } g_1 = \lambda v.(v,()), \, g_2 = \lambda (v,()).v,\\
        \qtabs{2} bx_v = \product{Replace}{(Skip \ (\lambda \_.())}$

The above program rearranges the source, a non-empty list, to a pair of its head element $s$ and its tail $ss$, and the view to a pair $(v, ())$, then we can use $v$ to replace $s$ and $()$ to keep $ss$. Intuitively, $\putbx{phead}{s_0}{v_0}$ returns a list whose head is $v_0$ and tail is the tail of $s_0$, and $\getbx{phead}{s_0}$ returns the head of the list $s_0$. For instance, $\putbx{phead}{[1,2,3]}{100} = [100,2,3]$ and $\getbx{phead}{[1,2,3]} = 1$. If we want to update the head element of the head element of a list of lists by using the view, we can define a composition like $phead \circ phead$. For example:

@@

In the end of this section, let's see the minBiGUL definition of $bfoldr$ which is a putback function of an important higher-order function on lists, $foldr$:

@@

\noindent In $bfoldr$, the composition is inherently left associative, and the number of composition is dynamically determined by the length of the source list. Since $\circ$ has a higher priority than $\times$, it is seemingly impossible to transform $bfoldr$ from the left associative composition style to the right one. 
%will look like a composition of many programs at some points in the evaluation if we slow down this process. Since $\circ$ has a higher priority than $\times$, it is seemingly impossible to transform $bfoldr$ from the left associative composition style to the right one.
